# (PART\*) Decision making {-}

# Decision making

It is not rare that we wish to *conditionally* perform operations - or in other words, to perform different operations depending on the outcome of a test of desired conditions.

We can use `if() {}`, `if() {} else {}` and `ifelse()` to help us with this *selection* process.

# `if` **statement**

In the example below, the `expression` is only executed if the `condition` that is being tested is `TRUE`.

<div class = "split">
<div class = "uneven-split-left">

```{r, echo = TRUE, eval = FALSE}
if(condition) {
  expression
}
```

</div>
<div class = "uneven-split-right">
<center>
```{r flowchart2, echo=FALSE, purl=FALSE}
#Create node data frame
f2_nodes <- create_node_df(n = 4, #number of nodes
                           type = 'a', #for grouping
                           label = c(" ", "Condition", "Expression", " "), 
                           shape = c("circle", "diamond", 'rectangle', 'circle'), 
                           style = c('rounded', 'solid', 'solid', 'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
                           )


#create edges
f2_edges <- create_edge_df(from = c(1, 2, 2, 3), #origin node id
                           to = c(2, 3, 4, 4), 
                           label = c(' ', '  if TRUE', '  if FALSE', ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 'w', 's'),
                           headport = c('n', 'n', 'n')
                           )

#create flowchart
flowchart2 <- create_graph(
                    nodes_df = f2_nodes,
                    edges_df = f2_edges,
                    directed = TRUE, 
                    attr_theme = 'tb' 
                    )

#display flowchart
render_graph(flowchart2, width = '45%', height = 'auto')
```
</center>
</div>
</div>

::: explanation

Note that when the `condition` is `FALSE`, nothing happens and the program carries on to the next step.

:::

# `if else` **statement**

The `if` statement only performs an operation if a condition is met. However, one may not only be interested in specifying the `expression` that must be done when the required `condition` is `TRUE`, but also the `expression` that must be performed when the `condition` is `FALSE`.

To achieve this, we can use the `if() {} else {}` statement and `ifelse()` function.

<div class = "split">
<div class = "uneven-split-left">

```{r, echo = TRUE, eval = FALSE, purl=FALSE}
if(condition) {
  expression 1
} else {
  expression 2
}
```

```{r, echo = TRUE, eval = FALSE, purl=FALSE}
ifelse(condition, 
       expression 1, 
       expression 2
)
```


</div>
<div class = "uneven-split-right">
<center>
```{r flowchart3, echo=FALSE, purl=FALSE}
#Create node data frame
f3_nodes <- create_node_df(n = 5, #number of nodes
                           type = 'a', #for grouping
                           label = c(" ", "Condition", "Expression 1\nif body", " ", "Expression 2\nelse body"), 
                           shape = c("circle", "diamond", 'rectangle', 'circle', 'rectangle'), 
                           style = c('rounded', 'solid', 'solid', 'filled', 'solid'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
                           )


#create edges
f3_edges <- create_edge_df(from = c(1, 2, 2, 3, 5), #origin node id
                           to = c(2, 3, 5, 4, 4), 
                           label = c(' ', ' if TRUE', '  if FALSE', ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 'w', 's', 's', 's'), #where edge leaves node
                           headport = c('n', 'n', 'n', 'n', 'n') #where arrow meets node
                           )

#create flowchart
flowchart3 <- create_graph(
                    nodes_df = f3_nodes,
                    edges_df = f3_edges,
                    directed = TRUE, 
                    attr_theme = 'tb' 
                    )

#display flowchart
render_graph(flowchart3, width = '50%', height = 'auto')
```
</center>
</div>
</div>

# Multiple condition decisions

While simple `if` and `if else` operations tests single conditions, we can make more complex operations, which tests multiple conditions.

With the `ifelse()` function, we can test a test a vector of conditions and/or apply functions only under certain conditions. See the examples below:

<div class = "split">
<div class = "split-left">

```{r, echo = c(-1) }
options(width = 30)
a <- 1:10

ifelse(test = a > 5, 
       yes = "yes", 
       no = "no")
```

</div>
<div class = "uneven-split-right">

```{r}
a <- (-4):5

sqrt(ifelse(test = a >= 0, 
            yes = a,
            no = NA)
     )
```

</div>
</div>


While the `if()` and `if()` `else` statements leave you with exactly two options, nested `if()` `else` statement allows you consider more alternatives.

<div class = "split">
<div class = "uneven-split-left">

```{r, eval = FALSE, purl=FALSE}
if(test_expression1) {
  statement1
} else if(test_expression2) {
  statement2
} else if(test_expression3) {
  statement3
} else {
  statement4
}
```

</div>
<div class = "uneven-split-right">
<center>

```{r flowchart4, echo=FALSE, purl=FALSE}
#note this one might turn out better in diagraph / DOT language, 'rank = same' seems broken without DOT
#Create node data frame
f4_nodes <- create_node_df(n = 9, #number of nodes
                           group = c('a','b','b','b','c','c','c','c','a'),
                           label = c("Start", "Condition 1", "Condition 2", "Condition 3", 
                                     "Expression 1", "Expression 2", "Expression 3", "Expression 4",  "End"), 
                           shape = c("circle", "diamond", "diamond", "diamond", 
                                     'rectangle', 'rectangle','rectangle', 'rectangle', 'circle'), 
                           style = c('solid', 'solid', 'solid', 'solid', 
                                     'solid', 'solid', 'solid', 'solid', 'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 12,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
                           )


#create edges
f4_edges <- create_edge_df(from = c(1, 2, 2, 3, 3, 4, 4, 5, 6, 7, 8), #origin node id
                           to = c(2, 3, 5, 4, 6, 7, 8, 9, 9, 9, 9), 
                           label = c(' ', '  if FALSE', '  if TRUE', ' if FALSE', '  if TRUE', ' if FALSE', ' if TRUE',
                                     ' ', ' ', ' ', ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           rel = c('a','b','c', 'b', 'c', 'c', 'd', 'd', 'd', 'd,'),
                           tailport = c('s', 'e', 's', 'e', 's', 'e', 's'), #where edge leaves node
                           headport = c('n', 'n', 'n', 'n', 'n', 'n', 'n') #where arrow meets node
                           )

#create flowchart
flowchart4 <- create_graph(
                    nodes_df = f4_nodes,
                    edges_df = f4_edges,
                    directed = TRUE,
                    attr_theme = 'tb'
                    )

#display flowchart
render_graph(flowchart4, width = '75%', height = 'auto')
```

</center>
</div>
</div>

# Caution with parsing statements

::: noway

What do you think will happen if we try the code below?

<div class = "split">
<div class = "split-left">

```{r bad-ifelse, eval = FALSE}
if(2+2) == 4 
print("Arithmetic works.")
else 
print("Houston, we have a problem.")
```

</div>
<div class = "split-right">

```{r bad-ifelse-out, echo = FALSE, error = TRUE}
if(2+2) == 4 
print("Arithmetic works.")
else
print("Houston, we have a problem.")
```

</div>
</div>

This does not work because `R` evaluates the first line and does not know that you are going to use an `else` statement
:::

::: explanation

Use curly brackets `{ }` so that `R` knows to expect more input. Try:

```{r}
if(2+2 == 4) {
  print("Arithmetic works.")
} else {
  print("Houston, we have a problem.")
}
```

:::

# Challenge 1

Welcome to our first challenge! To complete it, you will use the code block below, and you must:

<div class = "split">
<div class = "uneven-split-left">

1. Use an `if()` statement to print “meow” if `Paws` is a “cat”.
2. Use an `if()`  `else` statement to print `"woof"` if you supply an object that is a `"dog"` and `"meow"` if it is not. Try it out with `Paws` and `Scruffy`.
3. Use the `ifelse()` function to display `"woof"` for `animals` that are dogs and `"meow"` for `animals` that are cats.

</div>
<div class = "uneven-split-right">

```{r, echo = TRUE, eval = TRUE}
Paws <- "cat"
Scruffy <- "dog"
Sassy <- "cat"

animals <- c(Paws, Scruffy, Sassy)
```

</div>
</div>

::: explanation 

Here is a cheatsheet with logical operators that can help you completing this exercise: 

| Command | Meaning |
| :-------------: | :-------------: |
| `==`  | equal to  |
| `!=`  | not equal to  |
| `<`   | less than  |
| `<=`  | less than or equal to  |
| `>`   | greater than  |
| `>=`  | greater than or equal to  |
| `x&y` | `x` AND `y` |
| `x`&#124;`y`  | `x` OR `y`  |
| `isTRUE(x)`  | test if `x` is true |
:::

<script>
function myFunction() {
    var x = document.getElementById("myDIV");
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>
 
<button onclick="myFunction()">Solution for Challenge 1</button>
 
<div id="myDIV">

1. Use an `if()` statement to print “meow” if `Paws` is a “cat”.

```{r}
if(Paws == 'cat') {
  print("meow")
}
```

2\. Use an `if()`  `else` statement to print `“woof”` if you supply an object that is a `“dog”` and `“meow”` if it is not. Try it out with `Paws` and `Scruffy`.

```{r}
x = Paws
# x = Scruffy
if(x == 'cat') {
  print("meow")
} else {
  print("woof")
}
```


3\. Use the `ifelse()` function to display `“woof”` for `animals` that are dogs and `“meow”` for `animals` that are cats.

```{r}
animals <- c(Paws, Scruffy, Sassy)

ifelse(animals == 'dog', "woof", "meow")
```

or

```{r}
for(val in 1:3) {
  if(animals[val] == 'cat') {
    print("meow")
  } else if(animals[val] == 'dog') {
    print("woof")
  } else print("what?")
}
```

</div>
