# (PART\*) Iteration {-}

# Iteration

Every time some operations have to be repeated, a *loop* may come in handy. Loops are good for:

- Doing something for every element of an object;
- Doing something until the processed data runs out;
- Doing something for every file in a folder;
- Doing something that can fail, until it succeeds;
- Iterating a calculation until it reaches convergence.

Here, we will learn the following types of *iteration* statements:

```{r, echo = TRUE, eval = FALSE, purl=FALSE}
for() {}
while() {}
repeat {}
```

We will also dive into the `apply()` family functions, which are interesting function-based alternatives to `for() {}` loops.

We will finally learn about two special statements that allow us to control iterations:

```{r, echo = TRUE, eval = FALSE, purl=FALSE}
break
next
```

# `for` loops

A `for()` loop works in the following way:


```r
for(i in sequence) {
expression
}
```

<center>

```{r flowchart5for-loop, echo = FALSE, purl=FALSE}
#Create node data frame
f5_nodes <- create_node_df(n = 4, #number of nodes
                           type = 'a', #for grouping
                           label = c(" ", "Last item\nreached?", "expression\nfor() \nbody", "  "), 
                           shape = c("circle", "diamond", 'rectangle', 'circle'), 
                           style = c('solid', 'solid', 'solid', 'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f5_edges <- create_edge_df(from = c(1, 2, 2, 3), #origin node id
                           to = c(2, 3, 4, 2), 
                           label = c(' ', ' if FALSE', '  if TRUE, exit loop', ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 'w', 'e', 'e'),
                           headport = c('n', 'w', 'n', 's')
)

#create flowchart
flowchart5 <- create_graph(
  nodes_df = f5_nodes,
  edges_df = f5_edges,
  directed = TRUE, 
  attr_theme = 'tb' 
)

#display flowchart
render_graph(flowchart5, width = '35%', height = 'auto')
```

</center>

::: explanation

The letter `i` can be replaced with any variable name, `sequence` can be elements or the position of these elements, and `expression` can be anything. Try the examples below:

<div class = "split">
<div class = "split-left">

```{r, echo = TRUE, eval = FALSE}
for(a in c("Hello", 
           "R", 
           "Programmers")) {
  print(a)
}
```


```{r, echo = TRUE, eval = FALSE}
for(z in 1:4) {
  a <- rnorm(n = 1, 
             mean = 5 * z, 
             sd = 2)
  print(a)
}
```

</div>
<div class = "split-right">

<script>
function myFunction() {
var x = document.getElementById("myDIV");
if (x.style.display === "none") {
x.style.display = "block";
} else {
x.style.display = "none";
}
}
</script>

<button onclick="myFunction()">Output</button>

<div id="myDIV">

```{r, echo = FALSE, eval = TRUE, purl=FALSE}
for(a in c("Hello", 
           "R", 
           "Programmers")) {
  print(a)
}
```

<br>

```{r, echo = FALSE, eval = TRUE, purl=FALSE}
for(z in 1:4) {
  a <- rnorm(n = 1, 
             mean = 5 * z, 
             sd = 2)
  print(a)
}
```

</div>
</div>
</div>

:::

# Examples using `for`

## Simple `for` loop

In this next example, every instance of `m` is being replaced by each number between `1` and `7`, until it reaches the last element of the sequence:

<div class = "split">
<div class = "split-left">

```{r eval = TRUE, echo = TRUE}
y <- 2
for(m in 1:6) {
  print(y*m)
}
```

</div>
<div class = "split-right">

<center>
```{r flowchart-extrafor-loop, echo = FALSE, purl=FALSE}
#Create node data frame
f5_nodes <- create_node_df(n = 5, #number of nodes
                           type = 'a', #for grouping
                           label = c("Start", 
                                     "y = 2", 
                                     "for m in 1:6\nis m > 6?",
                                     "print(y*m)", 
                                     "Exit"), 
                           shape = c("circle", 
                                     "parallelogram", 
                                     "diamond",
                                     "parallelogram", 
                                     'circle'), 
                           style = c('solid', 
                                     'solid', 
                                     'solid', 
                                     'solid', 
                                     'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f5_edges <- create_edge_df(from = c(1, 2, 3, 3, 4), #origin node id
                           to = c(2, 3, 4, 5, 3), 
                           label = c(' ',
                                     ' ',
                                     ' if FALSE', 
                                     ' if TRUE', 
                                     ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 's', 'w', 'e', 'e'),
                           headport = c('n', 'n', 'w', 'n', 's')
)

#create flowchart
flowchart5 <- create_graph(
  nodes_df = f5_nodes,
  edges_df = f5_edges,
  directed = TRUE, 
  attr_theme = 'tb' 
)

#display flowchart
render_graph(flowchart5, width = '60%', height = 'auto')
```
</center>

</div>
</div>

## `for` loops on different classes

As expected, you can use `for()` loops in different object types and classes, such as a `list`. Let us take the example below, where we are creating the `elements` object list.

```{r echo = TRUE}
(elements <- list(a = 1:3, 
                  b = 4:10, 
                  c = 7:-1))
```

Now, let us `print` the double of every element of the list:

```{r echo = TRUE}
for(element in elements) {
  print(element*2)
}

```

## `for` and `if` together

Let us perform operations for even elements within `x` using the modulo operator (`%%`):

<div class = "split">
<div class = "split-left">

```{r, echo = TRUE, eval = TRUE}
x <- c(2, 5, 3, 9, 6)
count <- 0
```

</div>
<div class = "split-right">

```{r, echo = TRUE, eval = TRUE}
for(val in x) {
  if(val %% 2 == 0) {
    count <- count + 1
  }
}
print(count)
```

</div>
</div>

The above example can be represented within the following flowchart:


<center>
```{r flowchart6for-loop2, echo=FALSE, purl=FALSE}
#Create node data frame
f6_nodes <- create_node_df(n = 6, #number of nodes
                           type = 'a', #for grouping
                           label = c("Start",
                                     "x",
                                     "Last VAL\nin X?", 
                                     "Is VAL even?", 
                                     'count = count + 1',
                                     "print\ncount"), 
                           shape = c("circle",
                                     "parallelogram",
                                     "diamond", 
                                     'diamond', 
                                     'parallelogram', 
                                     'circle'), 
                           style = c('solid',
                                     'solid',
                                     'solid', 
                                     'solid', 
                                     'solid',
                                     'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f6_edges <- create_edge_df(from = c(1, 2, 3, 3, 4, 5, 4), #origin node id
                           to = c(2, 3, 4, 6, 5, 3, 3), 
                           label = c(' ', ' ', '  if FALSE', ' if TRUE', ' if TRUE', ' ', ' if FALSE'), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = FALSE,
                           tailport = c('e', 'e', 'e', 's', 'e', 'n', 'n'),
                           headport = c('w', 'w', 'w', 'n', 'w', 'n', 'n')
)

#create flowchart
flowchart6 <- create_graph(
  nodes_df = f6_nodes,
  edges_df = f6_edges,
  directed = TRUE, 
  attr_theme = 'lr' 
)

#display flowchart
render_graph(flowchart6, 
             width = '90%', 
             height = 'auto')
```
</center>


## `for` with a real dataset

`for()` loops are often used to loop over a dataset. We will use loops to perform functions on the `CO2` dataset which is built in `R`. To load and see the first 6 rows of the `CO2` dataset, execute the following code:

```{r, echo = TRUE, eval = TRUE}
data(CO2) # This loads the built in dataset

head(CO2)
```

Now, to recursively print the CO2 concentration, let us do this: 

```{r, echo = TRUE, eval = FALSE}
for(i in 1:length(CO2[,1])) { # for each row in the CO2 dataset
  print(CO2$conc[i]) # print the CO2 concentration
}
```


Here are the first 40 outputs:

<div class = "split">
<div class = "split-left">

<div class = "split">
<div class = "split-left">

```{r, echo=FALSE, purl=FALSE}
for(i in 1:10) { 
  print(CO2$conc[i]) 
}
```
</div>

<div class = "split-right">
```{r, echo=FALSE, purl=FALSE}
for (i in 11:20) { 
  print(CO2$conc[i]) 
}
```
</div>

</div>
</div>

<div class = "split-right">

<div class = "split">
<div class = "split-left">
```{r, echo=FALSE, purl=FALSE}
for (i in 21:30) { 
  print(CO2$conc[i]) 
}
```
</div>

<div class = "split-right">
```{r, echo=FALSE, purl=FALSE}
for (i in 31:40) { 
  print(CO2$conc[i]) 
}
```

</div>
</div>
</div>
</div>
</div>

Now, let us obtain the CO2 concentration only for the sites that were sampled in Québec (Canada):

```{r eval = FALSE, purl=FALSE}
for(i in 1:length(CO2[,1])) { # for each row in the CO2 dataset
  if(CO2$Type[i] == "Quebec") { # if the type is "Quebec"
    print(CO2$conc[i]) # print the CO2 concentration
  }
}
```


```{r, echo = TRUE, eval = FALSE, purl=FALSE}
for(i in 1:length(CO2[,1])) { # for each row in the CO2 dataset
  if(CO2$Type[i] == "Quebec") { # if the type is "Quebec"
    print(CO2$conc[i]) # print the CO2 concentration
  }
}
```


Here they are:

<div class = "split">
<div class = "split-left">

<div class = "split">
<div class = "split-left">

```{r, echo=FALSE}
for (i in 1:11) { 
  if(CO2$Type[i] == "Quebec") { 
    print(CO2$conc[i])
  }
}
```
</div>

<div class = "split-right">
```{r, echo=FALSE}
for (i in 12:22) { 
  if(CO2$Type[i] == "Quebec") { 
    print(CO2$conc[i])
  }
}
```
</div>

</div>
</div>

<div class = "split-right">

<div class = "split">
<div class = "split-left">
```{r, echo=FALSE}
for (i in 23:33) { 
  if(CO2$Type[i] == "Quebec") { 
    print(CO2$conc[i])
  }
}
```
</div>

<div class = "split-right">
```{r, echo=FALSE}
for (i in 34:42) { 
  if(CO2$Type[i] == "Quebec") { 
    print(CO2$conc[i])
  }
}
```

</div>
</div>
</div>
</div>
</div>

::: explanation 

**Tip 1.** To loop over the number of rows of a data frame, we can use the function `nrow()`:

```{r, echo = TRUE, eval = FALSE}
for(i in 1:nrow(CO2)) {
  # for each row in
  # the CO2 dataset
  print(CO2$conc[i])
  # print the CO2
  # concentration
}
```

::: 


::: explanation 

**Tip 2.** To perform operations on the elements of one column, we can directly iterate over it.

```{r, echo = TRUE, eval = FALSE}
for(p in CO2$conc) {
  # for each element of
  # the column "conc" of
  # the CO2 df
  print(p)
  # print the p-th element
}
```

::: 

::: explanation 

**Tip 3.** The expression within the loop can be almost anything and is usually a compound statement containing many commands.

```{r, echo = TRUE, eval = FALSE}
for(i in 4:5) { # for i in 4 to 5
  print(colnames(CO2)[i])
  print(mean(CO2[,i])) # print the mean of that column from the CO2 dataset
}
```

::: 

# `for` loops within `for` loops

In some cases, you may want to use nested loops to accomplish a task. When using nested loops, it is important to use different variables for each one of your loops. 

In the example below, we used a nested `for` framework which printed the product of `i` and `n`, where `n` was part of the nested `for` and was a sequence from `1` to `3`, and then `i`, which was also a sequence from `1` to `3`, was part of the outer loop.

<div class = "split">
<div class = "split-left">

```{r, echo = TRUE, eval = FALSE}
for (i in 1:3) {
  for (n in 1:3) {
    print (i*n)
  }
}
```

</div>

<div class = "split-right">

```{r, echo = FALSE, eval = TRUE}
for (i in 1:3) {
  for (n in 1:3) {
    print (i*n)
  }
}
```

</div>
</div>

::: explanation

Specifically, the operation above beings as follows:

1. `i` takes the *first* value from `1` to `3`, *i.e*. $i = 1$;
2. `n` takes the *first* value from `1` to `3`, *i.e*. $n = 1$;
3. The product of the current value of `i` and `n` is computed and printed, *i.e.* $i \times n = 1 \times 1 = 1$;
4. `n` takes the second value from `1` to `3`, *i*.*e*. $n = 2$;
5. The product of the current value of `i` and `n` is computed and printed, *i.e.* $i \times n = 1 \times 2 = 2$;
6. `n` takes the *third* value from `1` to `3`, *i*.*e*. $n = 3$;
7. The product of the current value of `i` and `n` is computed and printed, *i.e.* $i \times n = 1 \times 3 = 3$;
8. `i` takes the *second* value from `1` to `3`, *i.e*. $i = 2$;
9. The product of the current value of `i` and `n` is computed and printed, *i.e.* $i \times n = 2 \times 1 = 2$;
10. `n` takes the second value from `1` to `3`, *i*.*e*. $n = 2$;
11. The product of the current value of `i` and `n` is computed and printed, *i.e.* $i \times n = 2 \times 2 = 4$;
12. `n` takes the *third* value from `1` to `3`, *i*.*e*. $n = 3$;
13. The product of the current value of `i` and `n` is computed and printed, *i.e.* $i \times n = 2 \times 3 = 6$;
14. *Would you be able to follow up with the rest of the iteration?*

:::

# The `apply()` family

`R` disposes of the `apply()` function family, which consists of iterative functions that aim at **minimizing your need to explicitly create loops**.

## `apply()`

Let us consider that we have a `height` matrix containing the height (in metres) that was taken from five individuals (in rows) at four different times (as columns). 

```{r, echo = TRUE, eval = TRUE}
(height <- matrix(runif(20, 1.5, 2),
                  nrow = 5,
                  ncol = 4))
```

We would like to obtain the average height at each time step.

One option is to use a `for() {}` loop to iterate from column `1` to `4`, use the function `mean()` to calculate the average of the values, and sequentially store the output value in a vector.

Alternatively, we can use the `apply()` function to set it to apply the `mean()` function to every column of the `height` matrix. See the example below:

```{r, echo = TRUE, eval = TRUE}
apply(X = height,
      MARGIN = 2,
      FUN = mean)
```

> The `apply()` function begins with three arguments main arguments: `X`, which will take a matrix or a data frame; `FUN`, which can be any function that will be applied to the `MARGIN`s of `X`; and `MARGIN` which will take `1` for row-wise computations, or `2` for column-wise computations. 

## `lapply()`

`lapply()` applies a function to every element of a `list`.

The output returned is also `list` (explaining the “`l`” in `lapply`) and has the same number of elements as the object passed to it.

```{r, echo = TRUE, eval = TRUE}
SimulatedData <- list(
  SimpleSequence = 1:4,
  Norm10 = rnorm(10),
  Norm20 = rnorm(20, 1),
  Norm100 = rnorm(100, 5)
)
```

```{r, echo = TRUE, eval = TRUE}
# Apply mean to each element of the list

lapply(X = SimulatedData, 
       FUN = mean)
```

> `lapply()` operations done in objects different from a `list` will be coerced to a `list` via `base::as.list()`.

## `sapply()`

`sapply()` is a 'wrapper' function for `lapply()`, but returns a simplified output as a `vector`, instead of a `list`.

```{r, echo = TRUE, eval = TRUE}
SimulatedData <- list(SimpleSequence = 1:4,
                      Norm10 = rnorm(10),
                      Norm20 = rnorm(20, 1),
                      Norm100 = rnorm(100, 5))

# Apply mean to each element of the list
sapply(SimulatedData, mean)
```

## `mapply()`

`mapply()` works as a multivariate version of `sapply()`.

It will apply a given function to the first element of each argument first, followed by the second element, and so on. For example:

```{r, echo = TRUE, eval = TRUE}
lilySeeds <- c(80, 65, 89, 23, 21)
poppySeeds <- c(20, 35, 11, 77, 79)

# Output
mapply(sum, lilySeeds, poppySeeds)
```

## `tapply()`

`tapply()` is used to apply a function over subsets of a vector.

It is primarily used when the dataset contains dataset contains different groups (*i.e.* levels/factors) and we want to apply a function to each of these groups.

```{r, echo = TRUE, eval = TRUE}
head(mtcars)

# get the mean hp by cylinder groups
tapply(mtcars$hp, mtcars$cyl, FUN = mean)
```

# Challenge 2

You have realized that your tool for measuring uptake was not calibrated properly at Quebec sites and all measurements are 2 units higher than they should be.

1. Use a loop to correct these measurements for all Québec sites.

2. Use a vectorisation-based method to calculate the mean $CO_2$ uptake in both areas.

> For this, you must load the $\text{CO}_{2}$ dataset using `data(CO2)`, and then use the object `CO2`.

<script>
function myFunction() {
var x = document.getElementById("myDIV");
if (x.style.display === "none") {
x.style.display = "block";
} else {
x.style.display = "none";
}
}
</script>

<button onclick="myFunction()">Solution for Challenge 2</button>

<div id="myDIV">
1.-Using `for` and `if` to correct the measurements:

```{r, echo = TRUE, eval = TRUE}
for (i in 1:length(CO2[,1])) {
  if(CO2$Type[i] == "Quebec") {
    CO2$uptake[i] <- CO2$uptake[i] - 2
  }
}
```

2.- Using `tapply()` to calculate the mean for each group:

```{r, echo = TRUE, eval = TRUE}
tapply(CO2$uptake, CO2$Type, mean)
```

</div>

# Modifying iterations: `break`, `next` and `while`

In `R`, a standard looping sequence may be modified with the `break`, `next` and `while` statements.

## `next`

`next` allows us to skip an iteration if a certain condition is met. It is used inside the loop and is usually joined by the `if` statement, as shown below: 

```{r, echo = TRUE, eval = FALSE}
for(val in x) {
  if(condition) { next }
  statement
}
```

<center>
```{r flowchart8next, echo=FALSE, purl=FALSE}
#Create node data frame
f8_nodes <- create_node_df(n = 5, #number of nodes
                           type = 'a', #for grouping
                           label = c("Start", "Reached last \nval in X?\n\nfor()", 
                                     "Is condition \nTRUE?\n\nif()",
                                     'statement\nbody', "End"), 
                           shape = c("circle", "diamond", 'diamond', 'rectangle', 'circle'), 
                           style = c('solid', 'solid', 'solid', 'solid', 'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f8_edges <- create_edge_df(from = c(1, 2, 3, 4, 2, 3), #origin node id
                           to = c(2, 3, 4, 2, 5, 2), 
                           label = c(' ', ' if FALSE', '  if FALSE', ' ', 'if TRUE, then exit', ' if TRUE, then next'), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = FALSE,
                           tailport = c('e', 'e', 'e', 's', 'n', 's'),
                           headport = c('w', 'w', 'nw', 's', 'w', 's')
)

#create flowchart
flowchart8 <- create_graph(
  nodes_df = f8_nodes,
  edges_df = f8_edges,
  directed = TRUE, 
  attr_theme = 'lr' 
)

#display flowchart
render_graph(flowchart8, width = '95%', height = 'auto')
```
</center>

## `break`

The `break` statement is used inside a loop to terminate and exit the iteration when a certain condition is met. 

```{r, echo = TRUE, eval = FALSE, purl=FALSE}
for(val in x) {
  if(condition) { break }
  statement
}
```

<center>
```{r flowchart7break, echo=FALSE, purl=FALSE}
#Create node data frame
f7_nodes <- create_node_df(n = 6, #number of nodes
                           type = 'a', #for grouping
                           label = c("Start", 
                                     "Reached last \nval in X?\n\nfor()", 
                                     "Is condition \nTRUE? \n\nif()",
                                     'statement\nbody', 
                                     "End", 
                                     " "), 
                           shape = c("circle", "diamond", 'diamond', 'rectangle', 'circle', 'circle'), 
                           style = c('solid', 'solid', 'solid', 'solid', 'filled', 'invis'), 
                           fontname = 'Helvetica',
                           fontsize = 13,
                           fixedsize = 'false',
                           color = 'mediumblue',
                           width = .1,
                           penwidth = 1.5
)


#create edges
f7_edges <- create_edge_df(from = c(1, 2, 3, 4, 2, 3), #origin node id
                           to = c(2, 3, 4, 2, 5, 5), 
                           label = c(' ', ' if FALSE\nfor x', '  if FALSE', ' ', 'if TRUE, then exit', ' if TRUE, then break', ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = FALSE,
                           tailport = c('e', 'e', 's', 'w', 'n', 'e'),
                           headport = c('w', 'w', 'n', 's', 'w', 'w')
)

#create flowchart
flowchart7 <- create_graph(
  nodes_df = f7_nodes,
  edges_df = f7_edges,
  directed = TRUE, 
  attr_theme = 'lr' 
)

#display flowchart
render_graph(flowchart7, width = '95%', height = 'auto')
```
</center>

# `repeat` loops

A `repeat` loop performs an operation until it is deliberately stopped.

One may either use `Escape` to halt it, or more adequately, use `break`.

```{r eval = FALSE}
repeat {
  expression
  if {
    condition
  } break
}
```

<center>
```{r flowchart-repeat-loop, echo = FALSE, purl=FALSE}
#Create node data frame
f5_nodes <- create_node_df(n = 4, #number of nodes
                           type = 'a', #for grouping
                           label = c(" ", 
                                     "Expression",
                                     "Condition\nif()",
                                     "Exit"), 
                           shape = c("circle",
                                     "rectangle", 
                                     'diamond', 
                                     'circle'), 
                           style = c('solid', 
                                     'solid', 
                                     'solid',
                                     'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f5_edges <- create_edge_df(from = c(1, 2, 3, 3), #origin node id
                           to = c(2, 3, 2, 4), 
                           label = c(' ', 
                                     ' ', 
                                     ' if FALSE\nrepeat', 
                                     ' if TRUE\nbreak'),
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 'e', 'w', 's'),
                           headport = c('n', 'e', 'w', 'n')
)

#create flowchart
flowchart5 <- create_graph(
  nodes_df = f5_nodes,
  edges_df = f5_edges,
  directed = TRUE, 
  attr_theme = 'tb' 
)

#display flowchart
render_graph(flowchart5, 
             width = '35%', 
             height = '40%')
```
</center>

::: explanation

If you attempt to run the following chunk of code using `repeat`, it will display the result until you press `Escape`.

```{r eval = FALSE}
repeat {
  print("Press 'Esc' to stop me!")
}
```

```{r eval = FALSE, purl=FALSE}
[1] "Press 'Esc' to stop me!"
[1] "Press 'Esc' to stop me!"
...
...
[1] "Press 'Esc' to stop me!"
```

:::

# `while` loops

Within the `while` loop, an expression happens **while** a condition is met.

```{r eval = FALSE, purl=FALSE}
while(condition){
  expression
}
```

<center>
```{r flowchart-while-loop, echo = FALSE, purl=FALSE}
#Create node data frame
f5_nodes <- create_node_df(n = 4, #number of nodes
                           type = 'a', #for grouping
                           label = c(" ", 
                                     "Condition\nbeing met?",
                                     "Expression",
                                     "Exit"), 
                           shape = c("circle",
                                     "diamond", 
                                     'rectangle', 
                                     'circle'), 
                           style = c('solid', 
                                     'solid', 
                                     'solid',
                                     'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f5_edges <- create_edge_df(from = c(1, 2, 3, 2), #origin node id
                           to = c(2, 3, 2, 4), 
                           label = c(' ', 
                                     ' while TRUE', 
                                     ' ', 
                                     ' if FALSE\nexit'),
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 's', 'w', 'e'),
                           headport = c('n', 'n', 'w', 'n')
)

#create flowchart
flowchart5 <- create_graph(
  nodes_df = f5_nodes,
  edges_df = f5_edges,
  directed = TRUE, 
  attr_theme = 'tb' 
)

#display flowchart
render_graph(flowchart5, 
             width = '80%', 
             height = 'auto')
```
</center>

# Examples with modifying statements

We can print $\text{CO}_{2}$ concentrations for *chilled* treatments and keep count of how many replications using `next`.

```{r eval = TRUE, echo = TRUE}
count <- 0

for(i in 1:nrow(CO2)) {
  if(CO2$Treatment[i] == "nonchilled") next
  # Skip to next iteration if treatment is nonchilled
  count <- count + 1
  #  print(CO2$conc[i]) # You can turn this on if you want to
}
print(count) # The count and print command were performed 42 times.
```

```{r}
sum(CO2$Treatment == "chilled")
```

Alternatively, we could have written the example above using `repeat` and `break`:

```{r, echo = TRUE, eval = TRUE}
count <- 0
i <- 0
repeat {
  i <- i + 1
  if (CO2$Treatment[i] == "nonchilled") next  # skip this loop
  count <- count + 1
  print(CO2$conc[i])
  if (i == nrow(CO2)) break     # stop looping
}
print(count)
```

Finally, one could have also done this using a `while` loop:

```{r, echo = TRUE, eval = TRUE}
i <- 0
count <- 0
while (i < nrow(CO2))
{
  i <- i + 1
  if (CO2$Treatment[i] == "nonchilled") next  # skip this loop
  count <- count + 1
  print(CO2$conc[i])
}

print(count)
```

# Challenge 3

You have realized that another of your tools was not working properly!

At **Mississippi** sites, **concentrations** less than 300 were measured correctly, but concentrations equal or higher than 300 were overestimated by 20 units!

Your *mission* is to use a loop to correct these measurements for all Mississippi sites.

::: puzzle

**Tip.** Make sure you reload the $\text{CO}_{2}$ data so that we are working with the raw data for the rest of the exercise:

```{r, echo = TRUE, eval = TRUE}
data(CO2)
```

:::

<script>
function myFunction() {
    var x = document.getElementById("myDIV");
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>
 
<button onclick="myFunction()">Solution for Challenge 3</button>
 
<div id="myDIV">

Challenge 3: Answer

```{r, echo = TRUE, eval = TRUE}
for (i in 1:nrow(CO2)) {
  if(CO2$Type[i] == "Mississippi") {
    if(CO2$conc[i] < 300) next
    CO2$conc[i] <- CO2$conc[i] - 20
  }
}
```

We could also have written it in this way, which is more concise and clearer.

```{r, echo = TRUE, eval = TRUE}
for (i in 1:nrow(CO2)) {
  if(CO2$Type[i] == "Mississippi" && CO2$conc[i] >= 300) {
    CO2$conc[i] <- CO2$conc[i] - 20
  }
}
```
</div>

# Edit a plot using `for` and `if`

Let us plot uptake *versus* concentration with points of different colors according to their *type* (Québec or Mississippi) and *treatment* (chilled or nonchilled).

```{r, echo = TRUE, eval = TRUE}
plot(x = CO2$conc, y = CO2$uptake, type = "n", cex.lab=1.4,
     xlab = "CO2 concentration", ylab = "CO2 uptake")
# Type "n" tells R to not actually plot the points.
for (i in 1:length(CO2[,1])) {
  if (CO2$Type[i] == "Quebec" & CO2$Treatment[i] == "nonchilled") {
    points(CO2$conc[i], CO2$uptake[i], col = "red")
  }
  if (CO2$Type[i] == "Quebec" & CO2$Treatment[i] == "chilled") {
    points(CO2$conc[i], CO2$uptake[i], col = "blue")
  }
  if (CO2$Type[i] == "Mississippi" & CO2$Treatment[i] == "nonchilled") {
    points(CO2$conc[i], CO2$uptake[i], col = "orange")
  }
  if (CO2$Type[i] == "Mississippi" & CO2$Treatment[i] == "chilled") {
    points(CO2$conc[i], CO2$uptake[i], col = "green")
  }
}
```

# Challenge 4

Generate a plot of comparing the CO2 concentration *versus* uptake where each plant is shown using a different point.

**Bonus points** for doing this challenge using nested loops!

<script>
function myFunction() {
    var x = document.getElementById("myDIV");
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>
 
<button onclick="myFunction()">Solution for Challenge 4</button>
 
<div id="myDIV">

Challenge 4: Answer

```{r, echo = TRUE, eval = FALSE}
plot(x = CO2$conc, 
     y = CO2$uptake, 
     type = "n", 
     cex.lab=1.4, 
     xlab = "CO2 concentration", 
     ylab = "CO2 uptake") 

# Type "n" tells R to not actually plot the points.

plants <- unique(CO2$Plant)

for (i in 1:nrow(CO2)){ 
  for (p in 1:length(plants)) { 
    if (CO2$Plant[i] == plants[p]) { 
      points(CO2$conc[i],
             CO2$uptake[i],
             col = p) 
    } 
  } 
}
```

```{r, echo = TRUE, eval = FALSE}
plot(x = CO2$conc, y = CO2$uptake, type = "n", cex.lab = 1.4,
     xlab = "CO2 Concentration", ylab = "CO2 Uptake")

plants <- unique(CO2$Plant)

for (i in 1:nrow(CO2)){
  for (p in 1:length(plants)) {
    if (CO2$Plant[i] == plants[p]) {
      points(CO2$conc[i], CO2$uptake[i], col = p)
    }
  }
}
```

</div>