# (PART\*) Programming {-}

# Writing functions


Imagine that we would like to rescale variables to the range of `0` to `1`.

```{r echo = TRUE}
# our.dataset has 4 variables 

our.dataset <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```


The equation for rescaling variables into the simplex (0-1) is:

$$x_{\text{new}} =  \frac{x_i - \text{min}(x)}{ \text{max}(x) - \text{min}(x)}$$
We could rescale these four variables to `0` and `1` by doing the following:

```{r echo = TRUE}
our.dataset$a <- (our.dataset$a - min(our.dataset$a, na.rm = TRUE)) / 
  (max(our.dataset$a, na.rm = TRUE) - min(our.dataset$a, na.rm = TRUE))
our.dataset$b <- (our.dataset$b - min(our.dataset$b, na.rm = TRUE)) / 
  (max(our.dataset$b, na.rm = TRUE) - min(our.dataset$a, na.rm = TRUE))
our.dataset$c <- (our.dataset$c - min(our.dataset$c, na.rm = TRUE)) / 
  (max(our.dataset$c, na.rm = TRUE) - min(our.dataset$c, na.rm = TRUE))
our.dataset$d <- (our.dataset$d - min(our.dataset$d, na.rm = TRUE)) / 
  (max(our.dataset$d, na.rm = TRUE) - min(our.dataset$d, na.rm = TRUE))
```


**What if our dataset had 31 variables?**

Repeating that equation and this chunk of code 31 times could become a tedious and inneficient process:

```{r eval = FALSE}
our.dataset$a <- (our.dataset$a - min(our.dataset$a, na.rm = TRUE)) / 
  (max(our.dataset$a, na.rm = TRUE) - min(our.dataset$a, na.rm = TRUE))
```


But, we can see that, except from the *input*, the code was practically the same among the variables

*The function here was deliberately hidden to not cause confusion among the participants*

```{r eval = FALSE}
# our
# secret
# hidden
# function
```

```{r eval = FALSE}
rescale01(our.dataset$a)
rescale01(our.dataset$b)
rescale01(our.dataset$c)
rescale01(our.dataset$d)
```



####  Why write functions? #### 

Much of the heavy lifting in R is done by functions. They are useful
for:

1- Performing a task repeatedly, but configurably;

2- Making your code more readable;

3- Make your code easier to modify and maintain;

4- Sharing code between different analyses;

5- Sharing code with other people;

6- Modifying R's built-in functionality.

#### What is a function?


```{r flowchart9, echo=FALSE}
library(DiagrammeR)
#Create node data frame
f9_nodes <- create_node_df(n = 5, #number of nodes
                           group = c('a', 'a', 'a', 'b', 'c'),
                           label = c("Argument 1", "Argument 2",
                                     "Argument 3", "Data processing\nFunction", 'Return value'), 
                           shape = c("none", "none", 'none', 'rectangle', 'none'), #shape of node
                           style = c('none', 'none', 'none', 'rounded', 'none'), #line style around nodes
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE, #whether node changes size based on label or is fixed
                           color = 'mediumblue'
                           )
#create edges
##DiagrammeR doesn't handle NA's in edge labels well, so NAs are ' ' here.
f9_edges <- create_edge_df(from = c(1, 2, 3, 4), #origin node id
                           to = c(4, 4, 4, 5),
                           style = c('solid', 'solid', 'solid', 'dashed'),
                           fontsize = 10,
                           color = 'dimgrey',
                           headport = c('w', 'w', 'w', 'w'),
                           tailport = c('e', 'e', 'e', 'e')
                           )
#create flowchart
flowchart9 <- create_graph(
                    nodes_df = f9_nodes,
                    edges_df = f9_edges,
                    directed = TRUE, #arrows vs straight lines
                    attr_theme = 'lr' #left to right specified
                    )
#display flowchart
render_graph(flowchart9, 
             width = '95%', 
             height = 'auto')
```






#### Syntax of a function

```{r, echo = TRUE, eval = FALSE}
function_name <- function(argument1, argument2, ...) {
  expression...  # What we want the function to do
  return(value)  # Optional
}
```


- `function_name` is the name of the function, and will be stored in the `R` environment as an object with this name;
- `argument`s take the defined values that can be used within the function;
- `body` contains the statements that define what the function does;
- `output` contains the returned value from the function. If `return()` is absent, then the last expression is returned. 



#### Arguments of a function

```{r, echo = TRUE, eval = FALSE}
function_name <- function(argument1, argument2, ...) {
  body  # What we want the function to do
  return(values)  # Optional
}
```

Arguments are the *input* values of your function and will have the information your function needs to be able to perform correctly.


A function can have between 0 and an infinity of arguments. See the following example:

```{r, echo = TRUE, eval = FALSE}
operations <- function(number1, number2, number3) {
  result <- (number1 + number2) * number3
  print(result)
}
operations(1, 2, 3)
```

## Challenge 5

Using what you learned previously on flow control, create a function
print_animal that takes an animal as argument and gives the following
results:

```{r, echo = TRUE, eval = FALSE}
print_animal <- function(animal) {
  if (animal == "dog") {
    print("woof")
  } else if (animal == "cat") {
    print("meow")
  }
}
Scruffy <- "dog"
Paws <- "cat"

print_animal(Scruffy)

print_animal(Paws)
```

++++ Challenge 5: Answer \|

```{r, echo = TRUE, eval = FALSE}
print_animal <- function(animal) {
  if (animal == "dog") {
    print("woof")
  } else if (animal == "cat") {
    print("meow")
  }
}
```

++++

## Default values in a function

Arguments can be provided with a **default value**, or even be optional.

**Default values** are useful when using a function with the same settings. The flexibility to depart from default values is still there, if needed.

```{r, echo = TRUE, eval = FALSE}
operations <- function(number1, number2, number3 = 3) {
  result <- (number1 + number2) * number3
  print(result)
}
operations(number1 = 1, number2 = 2, number3 = 3) 
# is equivalent to
operations(1, 2)
operations(1, 2, 2) # we can still change the value of number3 if needed
```

## The ellipsis argument: `...`


The special argument `...` allows you to pass arguments from other undefined functions, *i.e.* allowing for an indefinite number of arguments inputs. 


```{r echo=TRUE}
paste_anything_fun <- function(...) {
  arguments <- list(...)
  paste0(arguments)
}
```

```{r echo=TRUE}
paste_anything_fun("I", 
                   "want", 
                   "a break!")
```

```{r echo=TRUE}
percentages <- function(x, mult = 100, ...){
 percent <- round(x * mult, ...)
 paste(percent, "%", sep = "")
}
```


```{r echo=TRUE}
percentages(c(.543, .534, .466))
``` 

```{r echo=TRUE}
# ?round
percentages(c(.543, .534, .466), digits = 2)
``` 


The special argument `...` allows you to pass on arguments to another function used inside your function. Here we use `...` to pass on arguments to `plot()` and `points()`.

```{r, eval = FALSE}
plot.CO2 <- function(CO2, ...) {
  plot(x=CO2$conc, y = CO2$uptake, type = "n", ...)
  for(i in 1:length(CO2[,1])){
     if(CO2$Type[i] == "Quebec") {
       points(CO2$conc[i], CO2$uptake[i], col = "red", type = "p", ...)
     } else if(CO2$Type[i] == "Mississippi") {
       points(CO2$conc[i], CO2$uptake[i], col = "blue", type = "p", ...)
     }
  }
}
```


```{r, eval = FALSE}
plot.CO2(CO2, 
         cex.lab = 1.2,
         xlab = "CO2 concentration", 
         ylab = "CO2 uptake")
```

```{r, eval = FALSE}
plot.CO2(CO2, 
         cex.lab = 1.2, pch = 20,
         xlab = "CO2 concentration", 
         ylab = "CO2 uptake")
```

```{r, echo = FALSE, fig.height = 2.5, fig.width = 5}
plot.CO2 <- function(CO2, ...) {
  plot(x = CO2$conc, y = CO2$uptake, type = "n", ...)
  for(i in 1:length(CO2[,1])){
     if(CO2$Type[i] == "Quebec") {
       points(CO2$conc[i], CO2$uptake[i], col="red", type="p", ...)
     } else if(CO2$Type[i] == "Mississippi") {
       points(CO2$conc[i], CO2$uptake[i], col="blue", type="p", ...)
     }
  }
}
par(mfrow=c(1, 2), mar = c(4, 4, 1, 1))
plot.CO2(CO2, 
         cex.lab=1.2, 
         xlab="CO2 concentration", ylab="CO2 uptake")
plot.CO2(CO2, cex.lab=1.2, xlab="CO2 concentration", ylab="CO2 uptake", pch=20)
```

## Return values

The last expression evaluated in a `function` becomes the return value:


```{r, echo = TRUE, eval = FALSE}
myfun <- function(x) {
  if (x < 10) {
    0
  } else {
    10
  }
}

myfun(5)
myfun(15)
```

`function()` itself returns the last evaluated value even without
including `return()` function.

It can be useful to explicitly `return()` if the routine should end
early, jump out of the function and return a value.

```{r, echo = TRUE, eval = FALSE}
simplefun1 <- function(x) {
  if (x<0)
  return(x)
}
```

Functions can return only a single object (and text). But this is not a
limitation because you can return a list containing any number of
objects.

```{r, echo = TRUE, eval = FALSE}
simplefun2 <- function(x, y) {
  z <- x + y
  return(list("result" = z,
              "x" = x,
              "y" = y))
}
```

```{r, echo = TRUE, eval = FALSE}
simplefun2(1, 2)
```

## Challenge 5

Using what you have just learned on functions and control flow, create a
function named `bigsum` that takes two arguments `a` and `b` and:

1. Returns $0$ if the sum of `a` and `b` is strictly less than $50$;

2. Else, returns the sum of `a` and `b`.

++++ Challenge 5 Answer \|

Answer 1:

```{r, echo = TRUE, eval = FALSE}
bigsum <- function(a, b) {
  result <- a + b
  if (result < 50) {
    return(0)
  } else {
    return (result)
  }
}
```

Answer 2:

```{r, echo = TRUE, eval = FALSE}
bigsum <- function(a, b) {
  result <- a + b
  if (result < 50) {
  0
  } else {
  result
  }
}
```


## Accessibility of variables

It is essential to always keep in mind where your variables are, and whether they are defined and accessible:

1. Variables defined inside a function are not accessible outside from it!

2. Variables defined outside a function are accessible inside, and are not modified, even if they have the same name.

```{r, error = TRUE}
out_val <- 3 
vartest <- function() {
  in_val <- 4  
  print(in_val)
  print(out_val)
}
vartest()
```

```{r, error = TRUE}
in_val; out_val
```


```{r}
out_val_2 <- 3
vartest <- function(out_val_2) {
  print(out_val_2)
}
vartest(8)
```

```{r, error = TRUE}
out_val_2
```

*What happens in the function club, stays in the function club.*

```{r, error = TRUE}
var1 <- 3 
vartest <- function() {
  a <- 4      # 'a' is defined inside
  print(a)    # print 'a'
  print(var1) # print var1
}
a             # we cannot print 'a' as it exists only inside the function
vartest()     # calling vartest() will print a and var1
rm(var1)      # remove var1
vartest()     # calling the function again does not work anymore
```


Be very careful when creating variables inside a conditional statement
as the variable may never have been created and cause (sometimes
unperceptible) errors.

It is good practice to define variables outside the conditions and then
modify their values to avoid any problem

```{r, echo = TRUE, eval = FALSE}
a <- 3
if (a > 5) {
  b <- 2
}

a + b
```

If you had `b` already assigned in your environment, with a different
value, you could have had a *bigger* problem!

No error would have been shown and `a + b` would have meant another
thing!

### Good programming practices

Why should I care about programming practices?

- It makes your life easier;
- It helps you achieve greater readability and makes sharing and reusing your code a lot less painful;
- It helps reduce the time you will spend remembering and understanding your own code.

### Pay attention to the next tips!

**Keep a clean and nice code**

Proper indentation and spacing is the first step to get an easy to read

code:

-   Use spaces between and after you operators

`x>=1&x<=10` is more difficult to read then `x >= 1 & x <= 10`


-   Use consistently the same assignation operator. 

<- is often preferred. = is OK, but do not switch all the time between the two


-   Use brackets when using flow control statements:

    -   Inside brackets, indent by *at least* two spaces;
    -   Put closing brackets on a separate line, except when preceding an `else` statement.
    
    
-   Define each variable on its own line.


- Use `Cmd + I` or `Ctrl + I` in RStudio to indent the highlighted code automatically;

```{r, eval = FALSE}
if((a[x,y]>1.0)&(a[x,y]<2.0)){print("Between 1 and 2")}
```

```{r, eval = FALSE}
 if((a[x, y] > 1.0) & (a[x, y] < 2.0)){
    print("Between 1 and 2")
  }
```


This code is not spaced, and therefore hard to read. All brackets are
badly aligned, and it looks "messy".

```{r, eval= FALSE}
a<-4;b=3
if(a<b){
if(a==0)print("a zero")}else{
if(b==0){print("b zero")}else print(b)}
```


This looks more organized, no?

```{r, echo = TRUE, eval = FALSE}
a <- 4
b <- 3
if(a < b){
  if(a == 0) {
    print("a zero")
  }
} else {
  if(b == 0){
    print("b zero")
  } else {
    print(b)
  }
}
```

### Use functions to simplify your code

Write your own function:

1.- When portion of the code is repeated more than twice in your script;

2.- If only a part of the code changes and includes options for
different arguments.

This would also reduce the number of potential errors done by
copy-pasting, and the time needed to correct them.

Let uzs now modify the example from Challenge \#3 and suppose that all
$CO_2$ uptake from Mississipi plants was overestimated by 20 and
Quebec underestimated by 50. We could write this:

```{r, echo = TRUE, eval = FALSE}
for (i in 1:length(CO2[,1])) {
  if(CO2$Type[i] == "Mississippi") {
    CO2$conc[i] <- CO2$conc[i] - 20
  }
}
for (i in 1:length(CO2[,1])) {
  if(CO2$Type[i] == "Quebec") {
    CO2$conc[i] <- CO2$conc[i] + 50
  }
}
```

Or this:

```{r, echo = TRUE, eval = FALSE}
recalibrate <- function(CO2, type, bias){
  for (i in 1:nrow(CO2)) {
    if(CO2$Type[i] == type) {
      CO2$conc[i] <- CO2$conc[i] + bias
    }
  }
  return(CO2)
}
```

```{r, echo = TRUE, eval = FALSE}
newCO2 <- recalibrate(CO2, "Mississipi", -20)
newCO2 <- recalibrate(newCO2, "Quebec", +50)
```

### Use meaningful names for functions

Same function as before, but with vague names

```{r, echo = TRUE, eval = FALSE}
rc <- function(c, t, b) {
  for (i in 1:nrow(c)) {
    if(c$Type[i] == t) {
      c$uptake[i] <- c$uptake[i] + b
    }
  }
  return (c)
}
```


Whenever possible, avoid using names of existing `R` functions and variables to avoid confusion and conflits.


### Use comments

Add comment to describe what your code does, how to use its arguments or
a detailed step-by-step description the function.

```{r, echo = TRUE, eval = FALSE}
# Recalibrates the CO2 dataset by modifying the CO2 uptake concentration
# by a fixed amount depending on the region of sampling

# Arguments
# CO2: the CO2 dataset
# type: the type ("Mississippi" or "Quebec") that need to be recalibrated.
# bias: the amount to add or remove to the concentration uptake

recalibrate <- function(CO2, type, bias) {
  for (i in 1:nrow(CO2)) {
    if(CO2$Type[i] == type) {
      CO2$uptake[i] <- CO2$uptake[i] + bias
    }
  }
  return(CO2)
}

```
