# (PART\*) Programming {-}

# Writing functions

\*\* Why write functions? \*\*

Much of the heavy lifting in R is done by functions. They are useful
for:

\- Performing a task repeatedly, but configurably;

\- Making your code more readable;

\- Make your code easier to modify and maintain;

\- Sharing code between different analyses;

\- Sharing code with other people;

\- Modifying R's built-in functionality.

#### What is a function?

#### Syntax of a function

```{r, echo = TRUE, eval = FALSE}
function_name <- function(argument1, argument2, ...) {
  expression...  # What we want the function to do
  return(value)  # Optional
}
```

#### Arguments of a function

```{r, echo = TRUE, eval = FALSE}
function_name <- function(argument1, argument2, ...) {
  expression...
  return(value)
}
```

Arguments are the entry values of your function and will have the
information your function needs to be able to perform correctly.

A function can have between 0 and an infinity of arguments. See the
following example:

```{r, echo = TRUE, eval = FALSE}
operations <- function(number1, number2, number3) {
  result <- (number1 + number2) * number3
  print(result)
}
operations(1, 2, 3)
```

## Challenge 5

Using what you learned previously on flow control, create a function
print_animal that takes an animal as argument and gives the following
results:

```{r, echo = TRUE, eval = FALSE}
print_animal <- function(animal) {
  if (animal == "dog") {
    print("woof")
  } else if (animal == "cat") {
    print("meow")
  }
}
Scruffy <- "dog"
Paws <- "cat"

print_animal(Scruffy)

print_animal(Paws)
```

++++ Challenge 5: Answer \|

```{r, echo = TRUE, eval = FALSE}
print_animal <- function(animal) {
  if (animal == "dog") {
    print("woof")
  } else if (animal == "cat") {
    print("meow")
  }
}
```

++++

## Default values in a function

Arguments can also be optional and be provided with a default value.

This is useful when using a function with the same settings, but still
provides the flexibility to change its values, if needed.

```{r, echo = TRUE, eval = FALSE}
operations <- function(number1, number2, number3 = 3) {
  result <- (number1 + number2) * number3
  print(result)
}

operations(1, 2, 3) # is equivalent to
operations(1, 2)
operations(1, 2, 2) # we can still change the value of number3 if needed
```

## Argument `...`

The special argument `...` allows you to pass on arguments to another
function used inside your function. Here we use `...` to pass on
arguments to `plot()` and `points()`

```{r, echo = TRUE, eval = FALSE}
plot.CO2 <- function(CO2, ...) { #<<
  plot(x=CO2$conc, y=CO2$uptake, type="n", ...) #<<
  for (i in 1:length(CO2[,1])){
     if (CO2$Type[i] == "Quebec") {
       points(CO2$conc[i], CO2$uptake[i], col = "red", type = "p", ...) #<<
     } else if (CO2$Type[i] == "Mississippi") {
       points(CO2$conc[i], CO2$uptake[i], col = "blue", type = "p", ...) #<<
     }
  }
}
plot.CO2(CO2, cex.lab=1.2, xlab="CO2 concentration", ylab="CO2 uptake") #<<
plot.CO2(CO2, cex.lab=1.2, xlab="CO2 concentration", ylab="CO2 uptake", #<<
         pch=20) #<<
```

The special argument `...` allows you to input an indefinite number of
arguments.

```{r, echo = TRUE, eval = FALSE}
sum2 <- function(...){ #<<
  args <- list(...) #<<
  result <- 0
  for (i in args)  {
    result <- result + i
  }
  return (result)
}

sum2(2, 3) #<<
sum2(2, 4, 5, 7688, 1) #<<
```

## Return values

The last expression evaluated in a `function` becomes the return value.

```{r, echo = TRUE, eval = FALSE}
myfun <- function(x) {
  if (x < 10) {
    0
  } else {
    10
  }
}

myfun(5)
myfun(15)
```

`function()` itself returns the last evaluated value even without
including `return()` function.

It can be useful to explicitly `return()` if the routine should end
early, jump out of the function and return a value.

```{r, echo = TRUE, eval = FALSE}
simplefun1 <- function(x) {
  if (x<0)
  return(x)
}
```

Functions can return only a single object (and text). But this is not a
limitation because you can return a list containing any number of
objects.

```{r, echo = TRUE, eval = FALSE}
simplefun2 <- function(x, y) {
  z <- x + y
  return(list("result" = z,
              "x" = x,
              "y" = y))
}
```

```{r, echo = TRUE, eval = FALSE}
simplefun2(1, 2)
```

## Challenge 6

Using what you have just learned on functions and control flow, create a
function named `bigsum` that takes two arguments `a` and `b` and:

Returns 0 if the sum of `a` and `b` is strictly less than 50; Else,
returns the sum of `a` and `b`

++++ Challenge 6 Answer \|

Answer 1:

```{r, echo = TRUE, eval = FALSE}
bigsum <- function(a, b) {
  result <- a + b
  if (result < 50) {
    return(0)
  } else {
    return (result)
  }
}
```

Answer 2:

```{r, echo = TRUE, eval = FALSE}
bigsum <- function(a, b) {
  result <- a + b
  if (result < 50) {
  0
  } else {
  result
  }
}
```


## Accessibility of variables

It is essential to always keep in mind where your variables are, and
whether they are defined and accessible:

Variables defined \*inside\* a function are not accessible outside of
it!

Variables defined \*outside\* a function are accessible inside. But it
is NEVER a good idea, as your function will not function if the outside
variable is erased.

```{r, echo = TRUE, eval = FALSE}
var1 <- 3     # var1 is defined outside our function
vartest <- function() {
  a <- 4      # 'a' is defined inside
  print(a)    # print 'a'
  print(var1) # print var1
}

a             # we cannot print 'a' as it exists only inside the function

vartest()     # calling vartest() will print a and var1

rm(var1)      # remove var1
vartest()     # calling the function again doesn't work anymore
```

Use arguments then!

Also, inside a function, arguments names will take over other variable
names.

```{r, echo = TRUE, eval = FALSE}
var1 <- 3     # var1 is defined outside our function
vartest <- function(var1) {
  print(var1) # print var1
}

vartest(8)    # Inside our function var1 is now our argument and takes its value

var1
```

Be very careful when creating variables inside a conditionnal statement
as the variable may never have been created and cause (sometimes
unperceptible) errors.

It is good practice to define variables outside the conditions and then
modify their values to avoid any problem

```{r, echo = TRUE, eval = FALSE}
a <- 3
if (a > 5) {
  b <- 2
}

a + b
```

If you had `b` already assigned in your environment, with a different
value, you could have had a \*bigger\* problem!

No error would have been shown and `a + b` would have meant another
thing!

### Good programming practices

Why should I care about programming practices?

\- To make your life easier;

\- To achieve greater readability and makes sharing and reusing your
code a lot less painful;

\- To reduce the time you will spend to understand your code.

### Pay attention to the next tips!

\*\* Keep a clean and nice code \*\*

Proper indentation and spacing is the first step to get an easy to read
code:

-   Use spaces between and after you operators
-   Use consistently the same assignation operator. \<- is often
    preferred. = is OK, but do not switch all the time between the two
-   Use brackets when using flow control statements:
    -   Inside brackets, indent by \*at least\* two spaces;
    -   Put closing brackets on a separate line, except when preceding
        an \`else\` statement.
-   Define each variable on its own line.

This code is not spaced, and therefore hard to read. All brackets are
badly aligned, and it looks \"messy\".

```{r, echo = TRUE, eval = FALSE}
a<-4;b=3
if(a<b){
if(a==0)print("a zero")}else{
if(b==0){print("b zero")}else print(b)
```

This looks more organized, no?

```{r, echo = TRUE, eval = FALSE}
a <- 4
b <- 3
if(a < b){
  if(a == 0) {
    print("a zero")
  }
} else {
  if(b == 0){
    print("b zero")
  } else {
    print(b)
  }
}
```

### Use functions to simplify your code

Write your own function:

1.- When portion of the code is repeated more than twice in your script;

2.- If only a part of the code changes and includes options for
different arguments.

This would also reduce the number of potential errors done by
copy-pasting, and the time needed to correct them.

------------------------------------------------------------------------

Let's modify the example from Challenge \#3 and suppose that all
\$CO_2\$ uptake from Mississipi plants was overestimated by 20 and
Quebec underestimated by 50. We could write this:

```{r, echo = TRUE, eval = FALSE}
for (i in 1:length(CO2[,1])) {
  if(CO2$Type[i] == "Mississippi") {
    CO2$conc[i] <- CO2$conc[i] - 20
  }
}
for (i in 1:length(CO2[,1])) {
  if(CO2$Type[i] == "Quebec") {
    CO2$conc[i] <- CO2$conc[i] + 50
  }
}
```

Or this:

```{r, echo = TRUE, eval = FALSE}
recalibrate <- function(CO2, type, bias){
  for (i in 1:nrow(CO2)) {
    if(CO2$Type[i] == type) {
      CO2$conc[i] <- CO2$conc[i] + bias
    }
  }
  return(CO2)
}
```

```{r, echo = TRUE, eval = FALSE}
newCO2 <- recalibrate(CO2, "Mississipi", -20)
newCO2 <- recalibrate(newCO2, "Quebec", +50)
```

### Use meaningful names for functions

Same function as before, but with vague names

```{r, echo = TRUE, eval = FALSE}
rc <- function(c, t, b) {
  for (i in 1:nrow(c)) {
    if(c$Type[i] == t) {
      c$uptake[i] <- c$uptake[i] + b
    }
  }
  return (c)
}
```

### Use comments

Add comment to describe what your code does, how to use its arguments or
a detailed step-by-step description the function.

```{r, echo = TRUE, eval = FALSE}
# Recalibrates the CO2 dataset by modifying the CO2 uptake concentration
# by a fixed amount depending on the region of sampling

# Arguments
# CO2: the CO2 dataset
# type: the type ("Mississippi" or "Quebec") that need to be recalibrated.
# bias: the amount to add or remove to the concentration uptake

recalibrate <- function(CO2, type, bias) {
  for (i in 1:nrow(CO2)) {
    if(CO2$Type[i] == type) {
      CO2$uptake[i] <- CO2$uptake[i] + bias
    }
  }
  return(CO2)
}

```

### Thank you for attending this workshop!